BIOL 274 Homework 8

### 2024-04-06 bio_dynamics/bio274_hw8.Rmd
finished 274 hw8

---
title: "BIOL 274 Homework 8"
author: "Ruby Krasnow"
date: "2024-04-05"
output:
  pdf_document:
    extra_dependencies: ["derivative", "tcolorbox", "xcolor"]
---
# Visualizing Solutions of Linear Systems and Runge-Kutta

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(phaseR)
library(demodelr)
```

@@

We then evaluate the Jacobian at our equilibrium point:
$$J(0,0)= \begin{bmatrix} 
    -0.025(0)+0.1 &  -0.025(0) \\
    0.025(0) & 0.025(0)-0.5 \end{bmatrix}= \begin{bmatrix} 
    0.1 &  0 \\
   0 & -0.5 \end{bmatrix} $$

We can see that our eigenvalues for this matrix are the values along the diagonal, $\lambda_1=0.1, \; \lambda_2=-0.5$. Using our knowledge of linear algebra, we know that the eigenvector associated with $\lambda_1$ is just $\begin{bmatrix}1 \\ 0 \end{bmatrix}$ and the eigenvector associated with $\lambda_2$ is $\begin{bmatrix}1 \\ 0 \end{bmatrix}$.
We have one positive and one negative eigenvalue, which means we will have an (unstable) saddle near the equilibrium at (0,0), with straight-line solutions tending towards the origin on the x-axis (because of the positive eigenvalue) and away from the origin on the y-axis (because of the negative eigenvalue).

```{r, echo=FALSE, message=FALSE, warning=FALSE,fig.keep="last"}
# Phase Portrait setup
y0 <- matrix(c(0, 1, 0, 4, -6, 1, 5, 0.5, 0, -3), 5, 2, byrow = TRUE)

y0_df <- tibble(x=c(0,0,-6,5,0), 
                y=c(1,4,1,0.5,-3),
                IC=c("V1", "V2", "V3", "V4", "V5")) %>% 
  mutate(pt = paste0("(",x, ",", y, ")"))

make_plane <- function(params) {
  params_unnamed <- unname(params)
  traj <- trajectory(fun_traj, y0 = y0, tlim= c(0, 1), tstep=0.005,
                     parameters = params_unnamed, add=FALSE)

@@

  A<- matrix(params_unnamed, 2, 2, byrow=TRUE)

  lambda1 <- eigen(A)$values[1] # eigenvalue 1
  lambda2 <- eigen(A)$values[2] # eigenvalue 2

  print(lambda1)
  print(lambda2)

  v1_col <- "gray"
  v1_wid <- 1
  v2_col <- "gray"
  v2_wid <- 1

  if (abs(lambda1) < (1^(-10))) {
    v1_col <- "black"
    v1_wid <- 2
    print("hi")
  }

  if (abs(lambda2) < (1^(-10))) {
    v2_col <- "black"
    v2_wid <- 2
    print("hi")
  }

  p1+theme_bw()+
    geom_vline(linewidth=v1_wid, xintercept = 0, color=v1_col)+
    geom_hline(linewidth=v2_wid, yintercept = 0, color=v2_col)+
    geom_path(data=z, aes(x=x, y=y, color=pt), linewidth=1.5)+
    geom_point(data=y0_df, aes(x=x,y=y, color=pt), shape=21, fill="white",size=4, stroke=2)+labs(color="IC")
}

@@

@@

@@

>>>>

---
title: "BIOL 274 Homework 8"
author: "Ruby Krasnow"
date: "2024-04-06"
output:
  pdf_document:
    extra_dependencies: ["derivative", "tcolorbox", "xcolor"]
---
# Visualizing Solutions of Linear Systems and Runge-Kutta

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(phaseR)
library(demodelr)
library(deSolve)
```

@@

We then evaluate the Jacobian at our other equilibrium point, (0,0):
$$J(0,0)= \begin{bmatrix} 
    -0.025(0)+0.1 &  -0.025(0) \\
    0.025(0) & 0.025(0)-0.5 \end{bmatrix}= \begin{bmatrix} 
    0.1 &  0 \\
   0 & -0.5 \end{bmatrix} $$

We can see that our eigenvalues for this matrix are the values along the diagonal, $\lambda_1=0.1, \; \lambda_2=-0.5$. Using our knowledge of linear algebra, we know that the eigenvector associated with $\lambda_1$ is just $\begin{bmatrix}1 \\ 0 \end{bmatrix}$ and the eigenvector associated with $\lambda_2$ is $\begin{bmatrix}1 \\ 0 \end{bmatrix}$.
We have one positive and one negative eigenvalue, which means we will have an (unstable) saddle near the equilibrium at (0,0), with straight-line solutions tending away from the origin on the x-axis (because of the positive eigenvalue) and towards the origin on the y-axis (because of the negative eigenvalue).

```{r, echo=FALSE, message=FALSE, warning=FALSE,fig.keep="last",fig.height=4, fig.width=5, fig.align='center'}
# Phase Portrait setup
y0 <- matrix(c(0.2, 1, 3, 4, -6, -1, -5, 1, 4, -3), 5, 2, byrow = TRUE)

y0_df <- tibble(x=c(0.2,3,-6,-5,4), 
                y=c(1,  4,-1,1,-3),
                IC=c("V1", "V2", "V3", "V4", "V5")) %>% 
  mutate(pt = paste0("(",x, ",", y, ")"))

make_plane <- function(params) {
  params_unnamed <- unname(params)
  traj <- trajectory(fun_traj, y0 = y0, tlim= c(0, 250), tstep=0.01,
                     parameters = params_unnamed, add=FALSE)

@@

  p1+theme_bw()+
    geom_vline(linewidth=1, xintercept = 0, color="gray")+
    geom_hline(linewidth=1, yintercept = 0, color="gray")+
    geom_path(data=z, aes(x=x, y=y, group=pt), linewidth=1.5)
}

@@

Given an initial value problem $\frac{dy}{dt} = f(t, y), \quad y(t_0) = y_0$, the Runge-Kutta method is as follows:

Pick a step-size $h>0$ and define:

\begin{align*}
y_{n+1} &= y_n + \frac{h}{6}\left(k_1 + 2k_2 + 2k_3 + k_4 \right),\\
t_{n+1} &= t_n + h \\
\end{align*}

for $n = 0, 1, 2, 3, ...$ using
\begin{align*}
 k_1 &= \ f(t_n, y_n), \\
 k_2 &= \ f\!\left(t_n + \frac{h}{2}, y_n + h \frac{k_1}{2}\right), \\ 
 k_3 &= \ f\!\left(t_n + \frac{h}{2}, y_n + h \frac{k_2}{2}\right), \\
 k_4 &= \ f\!\left(t_n + h, y_n + h k_3\right).
\end{align*}

We will first compute the result manually using this process, then check our results with two different R packages:
```{r}
manual<- tribble(
  ~val,  ~n_1,        ~n_2,         ~n_3,        ~n_4,
  "t",  0.5,        1,          1.5,        2,
  "k1", 0,         -0.63740945,-0.49940322,-0.28490054,
  "t_n+h/2",0.25,       0.75,       1.25,       1.75,
  "y_n+h*(k1)/2", 1,          0.6390269,  0.37485072, 0.23694178,
  "k2",-0.5,       -0.61253307,-0.35128265,-0.19649492,
  "y_n+h*(k2)/2", 0.875,      0.64524599, 0.41188086, 0.25904318,
  "k3",-0.3828125, -0.62451359,-0.42411461,-0.2348618,
  "y_n+h*(k3)", 0.80859375, 0.48612247, 0.28764422, 0.19073601,
  "k4",-0.65382385,-0.47263011,-0.24821759,-0.14552091,
  "y",  0.79837926, 0.49970152, 0.30816691, 0.20040567)

# from manual computation of intermediate values
as.data.frame(manual)
```

```{r}
# from demodelr package
demodelr_out<- as.data.frame(demodelr::rk4(system_eq=c(dy ~ -2*t*(y^2)),
              initial_condition = c(y=1),
              deltaT = 0.5,
              n_steps = 5))
demodelr_out

# from deSolve paackage
rk_fun <- function(t, x, parms) {
  with(as.list(parms), {
    dx <- -1*a*t*(x[1])^2
    list(dx) }) }

time <- seq(from=0, to=2, length.out=5)
deSolve_out <- as.data.frame(deSolve::rk4(c(y=1), time, rk_fun, c(a=2)))
deSolve_out
```

In all three cases, we found $y(2) \approx 0.2004057$.

@@

We define the total error after $k$ steps to be
$$e_k = \lvert y(t_k)-y_k\rvert $$
Since we know the true solution to our ODE is $y(t)=\frac{1}{1+t^2}$, we know that the value of the function at $t=2$ should be $y(2)=\frac{1}{1+4}=0.2$

```{r}
rk_est <- deSolve_out %>% filter(time==2) %>% pull(y) #our rk4 approx.
abs(0.2-rk_est) #abs. difference between approx. and real values
```

Thus, RK4 with $\Delta t=0.5$ gives us $e_4 = 4.057 \times 10^{-4}$.

@@

In class, we found that RK4 with $\Delta t=0.2$ gives us $e_{10} = 1.095 \times 10^{-5}$ and with $\Delta t=0.1$ gives us $e_{20} = 6.541 \times 10^{-7}$.

We know that
\begin{equation}\label{eqn:C}
e_n \leq C \Delta t
\end{equation},

so given the error associated with two different time steps $t_1$ and $t_2$, we have the following relationship:

$$e_{n_1} \leq C(\Delta t_1)^4, \quad e_{n_2} \leq C(\Delta t_2)^4 \quad\Longrightarrow\quad 
\frac{e_{n_1}}{e_{n_2}} \leq \left(\frac{\Delta t_1}{\Delta t_2}\right)^4 
\quad\Longrightarrow\quad 
\left(\frac{e_{n_1}}{e_{n_2}}\right)^{1/4} \leq \frac{\Delta t_1}{\Delta t_2} $$
Which we can rearrange as
$$\Delta t_1  \geq \left(\frac{e_{n_1}}{e_{n_2}}\right)^{1/4} \Delta t_2 $$

Given an interval $0 \leq t \leq T$ and a known time step $\Delta t_1$, you need $n_1=\frac{T}{\Delta t_1}$ time steps, so if we know the error $e_{n_2}$ associated with a time step $t_2$, we can find the number of steps needed to approximate a target error $e_{n_1}$ with the formula:

$$n_1  \geq \left(\frac{e_{n_1}}{e_{n_2}}\right)^{-1/4} \frac{T}{\Delta t_2} $$

We can write a function to do this in R:
```{r}
find_n <- function(e1, e2, t2, T) {
  out <- ((e1/e2)^(-1/4))*T*(1/t2)
  out
}
```

```{r}
e4<- 0.0004056722 # The error we calculated in part (b) above
e10<-1.095e-5     # From class slides
e20<- 6.541e-7    # From class slides

# Checking that our function works
test4 <- paste(round(find_n(e1=e4, e2=e10, t2=0.2, T=2),2),
      round(find_n(e1=e4, e2=e20, t2=0.1, T=2),2), sep=",")

test10<- paste(round(find_n(e1=e10, e2=e4, t2=0.5, T=2),2),
      round(find_n(e1=e10, e2=e20, t2=0.1, T=2),2), sep=",")

test20 <- paste(round(find_n(e1=e20, e2=e10, t2=0.2, T=2),2),
     round(find_n(e1=e20, e2=e4, t2=0.5, T=2),2), sep=",")

out <- paste(test4, test10, test20, sep="\n")
cat(out)
```
```{r}
en <- 0.0001 # Target error

# Applying the function to our target error and all three known error/time step combinations
out2 <- paste(find_n(e1=en, e2=e10, t2=0.2, T=2), 
              find_n(e1=en, e2=e20, t2=0.1, T=2),
              find_n(e1=en, e2=e4, t2=0.5, T=2), sep="\n")
cat(out2)
```

Therefore, we would need 6 time steps to approximate our solution with an error of less than 0.001.

We can see that we obtain the same result by finding $C$ directly and plugging in our values to (\ref{eqn:C}):

```{r}
c1<- 0.0004056722/(0.5)^4 #0.006490755
c2 <- 1.095e-5/(0.2)^4 #0.00684375
c3 <- 6.541e-7/(0.1)^4 #0.006541
c<- mean(c(c1, c2, c3))
2/((en/c)^(1/4))
```
